---
title: "R qPCR Tutorial"
author: "David Colameo"
date: "16 02 2022"
output: html_document
editor_options: 
  chunk_output_type: console
---


# Installing Needed Packages

We will need several packages in this tutorial. The function pkgTest checks if the packages are already installed and if it is, it will be loaded on your environment. You can add or delete packages in the packages-vector if you want to reuse this section.

```{r}
pkgTest <- function(x)
  {
    if (!require(x,character.only = TRUE))
    {
      install.packages(x,dep=TRUE)
        if(!require(x,character.only = TRUE)) stop("Package not found")
    }
  }

packages <- c("readxl", "reshape2", "ggplot2", "ggpubr", "dplyr", "rstatix", "lme4", "lmerTest", "emmeans")


for(package in packages){
  pkgTest(package)
}


```



# How To Load Data into R

There are numerous ways of how to import datasets into R. You can prepare them in Excel and save them either as excel-files or csv-files.
Other types of files are often also supported (txt, SQL-databases).

Prepare your data into a nice format in excel and keep it simple. It is important to not put more than one table per sheet. There is no need to make manual adjustments on excel, since those are much easier to implement in R and take more often than not just one line of code.

Lets start with a dummy qPCR-dataset, where the wells are labelled using the Plate-setup from the BioRad-program and then the results exported as excel-files

```{r}

## You should know where you saved your excel-file. I like to create a folder per markdown and save all related files in the same folder. You can find your file easier, you can use R-Studio auto-complete functionality by first writing "./" and then tab (|->) key on your keyboard.

excelPath <- "./TaqMan_Cq_Data.xlsx"
data <- read_excel(excelPath)

##Lets convert the data-object into a data.frame-object
data <- as.data.frame(data)


##There is a lot of "junk" in the table, since I didnt modify it manually on excel and took it as it was exported from the BioRad-program
## Lets clean it up a bit and select only the relevant columns and omit rows that are not complete (with NAs) because of empty wells

cols <- c("Target", "Sample", "Biological Set Name", "Cq")
data <- data[,cols]
data <- data[complete.cases(data),]
#Lets have a look
head(data)

```


# Table manipulations

Now that we have a clean table, we can start to modify it and add information to the table. For example, the triplicates are all stacked upon each other, we want to have them all in a row with columns like replicate1, replicate2, replicate3. Also RH stands for right hemisphere, LH stands for left hemispehere. I know that we injected into the RH always an LNA and a negative control (NC) into the left hemisphere except for Mouse 35. Another information to add is, M33 to M36 are injected with LNA218, M37 to M40 are injected with LNA129. These kind of manipulations heavily depends on how you organize your labelling of experiments and it is something you might need to change from analysis to analysis. Ofc if you find a system that works for you, try to keep it that way all the time to build a routine.


### Lets add triplicates
This section can be reused easily all the time

```{r}

#Here we are creating a column with unique names except for the triplicates. 
data$Group <- paste(data$Target, data$Sample, data$`Biological Set Name`, sep=".")

unique(data$Group)

##Lets add a new column Rep (for replicate) and then we will add in here whether a row is replicate 1, 2 or 3
data$Rep = NA
for(g in unique(data$Group)){
  
  data[data$Group == g,]$Rep <- paste("Rep", rep(1:nrow(data[data$Group == g,])), sep="")
  
}

#lets check
head(data)

#Now lets transform our data from long format to wide format
data <- dcast(data, Target + Sample + `Biological Set Name` ~ Rep, value.var = "Cq")

head(data)

##Voilà, looks much nicer now. You will notice how we have also a Rep5-6. This is the case, because during the pipetting of the qPCR I thought I did a pipetting error in one of the conditions, I added 3 more replicates for that condition

data[complete.cases(data),]
rowMeans(data[complete.cases(data),4:6])
rowMeans(data[complete.cases(data),7:9])
rowMeans(data[complete.cases(data),4:9])
t.test(data[complete.cases(data),4:6], data[complete.cases(data),7:9])

## Since it looks like the values of the 6 replicates are all very similar, we can do a mean of all 6 replicates

##Now lets add a column meanCq with the means of all 3 replicates
data$meanCq <- rowMeans(data[,4:9], na.rm = T)
head(data)


```



## House-Keeping Gene

Now lets add the values of the house-keeping gene. In this case it is U6. For that, we will split our data-table into two, one part with the genes of interest and the other half with the U6-values. Then we will merge those two tables back together column-wise (in the wide format). For splitting tables, subset is a useful-function

```{r}

##All genes except U6
data.targets <- subset(data, Target != "U6")
##Only U6
data.u6 <- subset(data, Target == "U6")

#lets merge those two tables according to Sample and Biological Set Name Column
data <- merge(data.targets,data.u6 , suffixes = c(".tar", ".u6"), by=c("Sample", "Biological Set Name"))



## Lets perform the delta Cq normalisation; which is subtracting meanCq of the targets from the meanCq of U6. This is equivalent to performing a division since we are in the log2-scale (2^Cq1 / 2^Cq2 -> Cq1 - Cq2 -> 2^(Cq1-Cq2))

data$dCq <- data$meanCq.tar - data$meanCq.u6


# You can also add the fold-change if you want

data$Fold.dCq <- 2^ -(data$dCq)
head(data)
```



### Lets add experiment-specific additional information

This section is heavily dependent on your dataset and how you organize your information.

```{r}
## ifelse is a useful tool to add such information very efficiently. You can give a statement to the function and it will add a value to a new column if it is true or add another value if it isnt. This is equivalent as for-looping over every row and ask the question but much more efficient and faster.


## if LH -> NC otherwise LNA
data$LNA <- ifelse(data$`Biological Set Name` == "LH", "NC", "LNA")

## First group of mice
m1 <- c("M33", "M34", "M35", "M36")

## If these mice belong to this group AND the hemisphere is right, lets add LNA218, otherwise the previous value (otherwise you will overwrite the NC you added before)
data$LNA <- ifelse(data$Sample %in% m1 & data$`Biological Set Name` == "RH", "LNA218", data$LNA )

## Same here with LNA129
m2 <- c("M37", "M38", "M39", "M40")

data$LNA <- ifelse(data$Sample %in% m2 & data$`Biological Set Name` == "RH", "LNA129", data$LNA )

## Lets add a column to group the mice in either LNA218 or LNA129 injected mice
data$Coort <- ifelse(data$Sample %in% m1, "Coort1 (LNA218)", "Coort2 (LNA129)")

##Lets change the values for that one mouse that was injected differently
data$LNA <- ifelse(data$Sample =="M35" & data$`Biological Set Name` == "RH", "NC", data$LNA)
data$LNA <- ifelse(data$Sample =="M35" & data$`Biological Set Name` == "LH", "LNA218", data$LNA)

head(data)

```


## Plot the Data

There is a separate in-depth tutorial on how to plot data properly. Using those templates we can create following plot. We have the measured miRNA (miR129 or miR218 in the X-axis and -dCq in the y-axis (normalized to U6)) We can further divide the plot in two using the Coort-column as factor in facet_grid. Plus, we need a boxplot for the right and left hemisphere (LNA vs Negative Control Injection) which we can achieve using the grouping-aesthetic. 

```{r}

p <- ggplot(data, aes(x=Coort, y=-dCq, color=Sample, fill=LNA, group= interaction(LNA, Coort)))  + 
  geom_boxplot(alpha=0.3, aes(fill=LNA))+
  
  geom_point(aes(color=Sample), alpha=0.8, position = position_dodge(0.75), size=3) + facet_grid(. ~ Target.tar, scales="free")

p


```

### Exercise for the fast ones

  1. Create a different plot where we have the microRNA on the x-axis and the two plots are split according to Coort-factor.
          - Which of the two kind of plots make more sense?
          
  2. Create a plot with the fold-changes in the y-axis using your preferred arrangement in the x-axis.
          - Tipp: use facet_wrap instead of facet_grid. Can you find out why?


#Statistics

In this section, we are going to perform some statistical analysis. There are many ways of doing this and there will be a more in-depth seminar on just ANOVAs and linear mixed models.


## ANOVA Assumptions

First we need to consider which test we want to use. For this, we need to make sure, we are meeting the ANOVA-assumptions.

We have one factor for which we want to test the effect: Does a specific LNA inhibit the measured microRNA? This means, we are considering a one-way ANOVA.

The assumptions for an ANOVA are:
  1. Experimental errors (residuals) are normally distributed (-> data does not necessarily need to be normally distributed)
  2. Equal variances between treatments (sphericity assumption)
  3. Independence of samples
  

The first two assumptions we can test visually (QQ-Plots) and with corresponding tests (Shapiro-Wilk normality test). The third assumption is based on your experimental design. In most of our cases in the lab, we don't meet this assumption, since primary cultures are dependent from each other (genetically identical cells). Or in our case, we have repeated measurements within a mouse (hemisphere and levels of miR218/miR129). These factors are dependent from each other. We will take into account dependency into the model by including the mouse-ID into the model.


### Normality Test

We will test normality with the Shapirp-Wilk test. The null-hypothesis is that the data is normally distributed. A p-value < 0.05 means, we reject the null-hypothesis -> the data is most likely not normally distributed.

```{r}

## Lets perform some descriptive statistics (number of samples per group, mean and standard deviation)

data %>%
  ## here we split our data by group. You will need to change this bit to your experimental setup
  group_by(LNA, Target.tar, Coort) %>%
  
  ## check with help(get_summary_stats) for more extensive options of summaries
  get_summary_stats(c(dCq, Fold.dCq), type = "common")


## Here we perform the Shapiro-Wilk normality test by splitting the data in groups
data %>%
  group_by(LNA, Target.tar, Coort) %>%
  shapiro_test(dCq) %>%
 add_significance("p")


## Here we perform the Shapiro-Wilk normality test by splitting the data in groups
data %>%
  group_by(LNA, Target.tar, Coort) %>%
  ##Lets take the fold changes here instead of dCq
  shapiro_test(Fold.dCq) %>%
 add_significance("p")



ggqqplot(data, "dCq", ggtheme = theme_bw()) +
  facet_grid(LNA ~ Target.tar, labeller = "label_both") + labs(title="QQ-Plots of group-splitted dCq values")

## A QQ-plot probably does not make a lot of sense with n=4 per group. However, it still follows the expected trendline quite nicely.


```


## ANOVA Statistics
Technically, we are not allowed to use a classical ANOVA to model our data, since our experimental design does not meet the 3rd requirement. For educational purposes, we will implement both analyses. If we put together in the same model LNA effects and gene effect, then we are performing a two-way ANOVA. Since we are using a multiplicative model (denoted by the * sign instead of a + sign) we can check for interaction-effects between LNA and gene-expression. This will also allow us later to perform contrasts on all levels. 

## Classical ANOVA
```{r}
## This command performs the ANOVA analysis and stores it in the variable mod (you can name it as you want ofc)
mod <- aov(-dCq ~ LNA*Target.tar, data)
summary(mod)

#lets test the normality of the residuals of our model

shapiro.test(residuals(mod))
ggqqplot(residuals(mod))
```

### Contrasting between Condition

In order to contrast the differences between conditions, we will perform a Dunnett's Test (via the trt.vs.ctrl command) where we specify NC as the control group. We can also perform a Tukey's test, where we do pairwise comparisons across all combinations (less power since we are performing more tests)

```{r}
##Now the contrasts, LNA-effect within a target (made possible by | command)
emmeans(mod, trt.vs.ctrl ~ LNA | Target.tar, ref="NC")

## Or Tukeys pairwise comparison test

emmeans(mod, pairwise ~ LNA | Target.tar)

```

## Repeated Measure ANOVA

All we need to do to perform a repeated measure ANOVA is add the term Error(Sample) where sample is the ID for the mice. The function residual does not work here and we need to extract them manually as seen below:

Do you note a difference in the results?

```{r}
## The Error(Sample)-term denotes the within subject dependence and this is how you would implement a repeated measure ANOVA
mod <- aov(-dCq ~ LNA*Target.tar + Error(Sample), data)
summary(mod)

#lets test the normality of the residuals of our model
res <- proj(mod)                                                                  
res <- res[[3]][, "Residuals"]
shapiro.test(res)
ggqqplot(res)
```

### Same principle for the contrasts


```{r}
##Now the contrasts, LNA-effect within a target (made possible by | command)
emmeans(mod, trt.vs.ctrl ~ LNA | Target.tar, ref="NC")

## Or Tukeys pairwise comparison test

emmeans(mod, pairwise ~ LNA | Target.tar)

```



## Linear Mixed Model

We will have a separate seminar on linear mixed models and nesting structures later. This is just a short introduction to show you how it is done and how easy it is to implement in R. The random-effect term is determined by the (1|Sample)- term inside the lmer-function

```{r}
mod <- lmer(-dCq ~ LNA*Target.tar +  (1|Sample), data)
summary(mod)


##Now the contrasts, LNA-effect within a target (made possible by | command)
emmeans(mod, trt.vs.ctrl ~ LNA | Target.tar, ref="NC")

## Or Tukeys pairwise comparison test

emmeans(mod, pairwise ~ LNA | Target.tar)
```


## Perform Paired T-Tests

What if we want to perform paired t-tests instead of ANOVAs? That's how you would do it. Of course, we need to correct for multiple testing:

```{r}

data$dCq <- -1*data$dCq
t.tests <- data %>% 
  ## We group here according to miRNA-levels and Coort
  group_by(Target.tar, Coort) %>% 
  ## We perform the t-test here, where paired is set as TRUE
    t_test(dCq ~ LNA, paired = T) 

## Add multiple testing correction, there are many. You can check a few, but the results are all similar -> no comparison passes the multiple correction
t.tests$p.adj <-  p.adjust(t.tests$p, method="fdr")


t.tests <- add_significance(t.tests, p.col="p.adj")

t.tests

```

### Exercise

Perform an unpaired t-test -> does the data change? And why?



## Classical ddCq Approach

What if we want to perform a classical ddCQ analysis. Again, a ddCq is nothing else than a division in the log-transform scale, meaning we need to subtract a condition by a control. In our case, we first normalized to a house-keeping gene and now we add a second layer of normalization, which is the LNA-effect vs negative control. Again, we can split the data into two and then re-merge them:


```{r}

## %in% is used if we have a condition with more than one matching condition (instead of concatenating twice the same statement)
lna <- subset(data, LNA %in% c("LNA218", "LNA129"))
ctrl <- subset(data, LNA == "NC")

## We will merge the two tables according to mouse and coort, which will make sure that we are pairing the right negative controls with the right condition

ddCq.data <- merge(lna, ctrl, suffixes = c(".lna", ".ctrl"), by=c("Sample", "Coort", "Target.tar"))

head(ddCq.data)

## Now we have a rather wide data.frame. At this point we could also clip the data a bit and omit useless columns. Or we keep just going. We just need to add a normalized ddCq column

ddCq.data$ddCq <- ddCq.data$dCq.lna - ddCq.data$dCq.ctrl
ddCq.data$fold.ddCq <- 2^- ddCq.data$ddCq


```


## Lets Plot the ddCq

We will add a dashed line at the intercept for 0, indicating no change (log(0) = 1 because x^0 = 1)

```{r}

p <- ggplot(ddCq.data, aes(x=Target.tar, y=-ddCq, group= Target.tar))  + 
geom_boxplot(alpha=0.4, aes(fill=Target.tar))+
  
  geom_point(aes(color=Sample), alpha=0.8, position = position_dodge(0.75), size=3) + facet_grid(.~Coort, scales="free")




p+ geom_hline(yintercept = 0, linetype="dashed")


```

### Exercise

Create a plot with the fold-changes in the y-axis. Change also the intercept to an appropriate number.

## Statistics

With the knowledge


Similarly, we will perform an one-way ANOVA.

```{r}
mod <- aov(-ddCq ~ Coort*Target.tar + Error(Sample), ddCq.data)

summary(mod)

emmeans(mod, pairwise ~ Target.tar | Coort)
## Does it make sense to compare miR129 to miR218? not really, but depending on your experimental setup, this could make sense.

```


## One Sample T-test

We could also perform a one sample t-tests per condition checking whether the effect of introducing an LNA elicited a change. The mu will be 0, since our null-hypothesis is no change hence 0. 

What should mu be if we dont perform the t-test in the log2-scale?

```{r}
##One sample T-test


ddCq.data$ddCq <- -1 * ddCq.data$ddCq


one.sample.TTest <- ddCq.data %>% 
                  group_by(Coort, Target.tar) %>% 
                      ##Here we define mu = 0
                            t_test(ddCq ~ 1, mu=0)

## we need to adjust the p-values for multiple testing
one.sample.TTest$p.adjusted <- p.adjust(one.sample.TTest$p, method = "fdr")

one.sample.TTest <- add_significance(one.sample.TTest, p.col="p.adjusted")

one.sample.TTest

```


Did you notice how these results are EXACTLY the same as the paired two-sample t-test we performed earlier? Why do you think?


## Exercise for you

Try to write a r-markdown script with one of your qPCR-datasets and analyze it using this tutorial as a template. You can send it to me for corrections and suggestions





